Characterizing the Role of the Human Virome in Colorectal Cancer

### 2017-03-02 doc/manuscript.Rmd
Have not pushed in a while. Lots of changes, especially to the discussion.

---
title: Characterizing the Human Virome and Colorectal Cancer Link Using Machine Learning and Network Theory
author: 
- name: Geoffrey D Hannigan
  affiliation: 1
- name: Melissa B Duhaime
  affiliation: 2
- name: Mack T Ruffin IV
  affiliation: 3
- name: Charlie C Koumpouras
  affiliation: 1
- name: Patrick D Schloss
  affiliation: 1,\*
address:
  - code: 1
    address: Department of Microbiology & Immunology, University of Michigan, Ann Arbor, Michigan, 48108
  - code: 2
    address: Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, Michigan, 48108
  - code: 3
    address: Department of Family Medicine, University of Michigan, Ann Arbor, Michigan, 48108
  - code: \*
    address: To whom correspondence may be addressed.
output:
  md_document:
    variant: markdown
---

@@

Colorectal cancer is the second leading cause of cancer-related death in the United States and is a primary cause of morbidity and mortality throughout the world. Although the cause of colorectal cancer remains unclear, it has been strongly linked to gut bacterial communities. Viruses are another important component of the gut microbial community that have yet to be studied in colorectal cancer, despite their oncogenic potential. We evaluated the gut virome (virus community) role in colorectal cancer using a cohort of 90 human subjects with either healthy, precancerous, or cancerous colons. We utilized 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic methods to compare the virome's role to that of the bacterial community. We found that alpha and beta diversity metrics were insufficient for detecting an association between the virome and colorectal cancer, but virome-based classification models were highly associated with colorectal cancer. Bacteriophages, not eukaryotic viruses, made up the majority of the CRC-associated virome, suggesting the community was indirectly linked to colorectal cancer, modulating bacterial community structure and functionality. Phages with broader host ranges had a more significant role in cancer development. These results provide foundational evidence that phage communities are associated with colorectal cancer, and that broadly tropic phages play more significant roles. Because of its importance, the virome should be considered in future colorectal cancer studies, as well as other cancer types.

@@

# Significance Statement
Colorectal cancer is a leading cause of cancer-related death in the United States and worldwide. It's progression and severity is linked to gut bacterial communities. Little is known about the cancer-associated gut virus communities and their influence on bacteria. We characterize this influence of gut virus community structure on colorectal cancer in humans. The link between the gut virome and colorectal cancer was established by using virome-based machine learning models to accurately classify patient cancer status. The cancer-virus link was driven primarily by bacteriophages (bacterial viruses). Wider phage host ranges were more strongly linked to colorectal cancer. The results suggest an indirect role for the virome impacting colorectal cancer by modulating their associated bacterial community.

@@

Development of colorectal cancer is a stepwise process that begins when healthy tissue develops into a precancerous polyp (i.e. adenoma) in the large intestine [@Fearon:2011go]. If left untreated, the adenoma will develop into a cancerous lesion that can invade and metastasize, leading to severe illness and death. Progression to cancer can be prevented when precancerous adenomas are detected and removed during routine screening [@Levin:2008kz; @Zauber:2015cr]. Survival for colorectal cancer patients may exceed 90% when the lesions are detected early and removed [@Levin:2008kz]. Screening methods are effective, but their invasiveness has created a lack of compliance, creating a need for accurate, non-invasive screening methods. One such method is screening associated gut microbial communities.

Although the cause of colorectal cancer remains unclear, it has been strongly associated with gut bacterial communities [@Flynn:2016iu; @Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb]. This association has allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja]. While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and understanding underlying etiology, bacteria are only a subset of the gut microbiome. Viruses are another important component of the gut microbial community that have yet to be studied in the context of colorectal cancer.

Due to their mutagenic abilities and their propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Additionally, because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk], and because bacteria have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb], bacteriophages are potentially indirectly linked to cancer. The gut virome (the virus community of the gut) has the potential to impact health and disease (e.g. cancer), and has been associated with diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb]. We aim to take this field of research further by beginning to assess the role of the virome in colorectal cancer.

@@

## Virome-Based Machine Learning Provides Accurate Cancer Classification
Previous work has shown that 16S rRNA community signatures are effective for classifying stool samples as originating from healthy, precancerous, or cancerous individuals [@Zackular:2014fba; @Baxter:2016dja]. This is valuable because it presents a potential alternative screening approach to an invasive and expensive colonoscopy. The exceptional performance of bacterial signatures in these predictive models also supports a role for bacteria in colorectal cancer. Here we built off of these findings by evaluating the ability of virus community signatures to classify stool samples and compared performance to models built using bacterial community signatures.

@@

We compared the bacterial 16S rRNA gene model to a model built using the virome signatures. The viral model performed significantly slightly better than the bacterial model (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) with the viral and bacterial models achieving mean AUC (area under the curve) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = sigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = sigfig)`, respectively, after fifteen random forest iterations **(Figure \ref{predmodel} A - B)**. To confirm that this observation was due to the virome signature itself, and was not a trait of metagenomic datasets in general, we built a model using whole metagenomic community signatures. This model performed poorly with a mean AUC of `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = sigfig)`, lending support to our observation that classification performance is a trait unique to the virus metagenome **(Figure \ref{predmodel} A - B)**.

@@

## The Gut Virome Classifies Cancerous, Pre-Cancerous, and Healthy States
After evaluating our ability to classify samples as cancerous or healthy, we incorporated the precancerous adenoma samples into the model and evaluated our ability to classify all three states in our total sample set. We used three-class random forest models for the bacterial 16S and viral sample sets. The bacterial signature model yielded an AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = sigfig)` and outperformed the viral community model which yielded an overall AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = sigfig)` (p-value = `r signif(threesig, digits = sigfig)`, **Figure \ref{threewaymodel} A-C**). Both models were best able to classify cancer samples from healthy or precancerous samples, but struggled to distinguish precancerous from healthy or cancerous **(Figure \ref{threewaymodel} A-B)**. The cancerous signal was the most discriminatory of the three sample types.

@@

## Bacteriophages Drive Link Between Virome and Colorectal Cancer
The virome-based model was able to accurately classify stool samples as cancerous, precancerous, or healthy. Not only is this important for establishing an alternative diagnostic model, but it also suggests an underlying biological importance for viruses in colorectal cancer. Above we used our classification models to evaluate which virus OGUs were most highly linked to colorectal cancer. We were able to further characterize these important OGUs to better understand the underlying biological link between the virome and colorectal cancer.

The role of the virome in colorectal cancer could have been driven directly by eukaryotic viruses or indirectly by bacteriophages acting through their bacterial hosts. To better understand the types of viruses that are important for colorectal cancer, we utilized the longest sequences from the OGUs as the representative sequences to be used for taxonomic classification. These sequences were aligned to a set of all reference virus genomes, including bacteriophages and eukaryotic viruses. A strict e-value threshold of 1e-25 was used to improve our confidence in the matches between the genome sequences. The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel} E)**. Overall we were able to identify `r signif(idcount, digits = sigfig)`% of the OGUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OGUs aligned to bacteriophage reference genomes. Thus the majority of the OGUs are bacteriophages and not eukaryotic viruses, indicating the association between the virome and colorectal cancer is reliant on bacteriophage communities. This is consistent with previous reports suggesting the gut virome is primarily phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb].

## Broadly Infectious Phages Play Greater Role in Colorectal Cancer
Reference-based analyses of the virome perform poorly due to sparse reference databases and genomic modularity. Much of the virome is described as genomic "dark matter" and taxonomic identities can be largely uninformative (e.g. *Siphoviridae* describes the phage morphology under an electron microscope). Sometimes these identities are also used to infer bacterial hosts. Instead of relying on genome alignments to infer the identities and host ranges of the phage OGUs, we employed a network-based technique to understand the roles of the OGUs in the greater microbiome context, as previously described (cite other network preprint here). We implemented a random forest model to predict which bacterial OGUs are infected by which phage OGUs, resulting in an ecological network of the bacteria and phages within the community **(Figure \ref{network} A)**. We calculated the alpha centrality (measure of importance in the ecosystem network) of each phage OGU's connection to the rest of the network, and compared the centrality to the importance of each OGU in the colorectal cancer classification model. We found that phage OGU centrality is significantly positively correlated with importance to the disease model (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, rho = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), indicating that phage network hubs may play more important roles in colorectal cancer **(Figure \ref{network} B)**.

# Discussion
Because of their propensity for mutagenesis and capacity for modulating their host functionality, many viruses are oncogenic [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Because some bacteria also have oncogenic properties, bacteriophages may play an indirect role in promoting carcinogenesis by altering bacterial communities [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb]. Despite their carcinogenic potential, the link between virus gut communities (the human gut virome) and cancer has yet to be evaluated. Here we show that, like gut bacterial communities, the gut virome is associated with colorectal cancer. Our findings support a model for oncogenesis by phage-modulated bacterial community composition.

Previous work has supported a causative role for bacterial community composition and colorectal cancer progression, as well as an indirect role for those antibiotic agents which alter that composition [@Zackular:2016en]. Zackular *et al* showed that administering antibiotics (*A*) and thereby killing subsets of bacterial communities (*B*) was sufficient for altering colorectal cancer progression (*C*) in an inflammation-based murine model for colorectal cancer. These findings established a model of cancer dependence on altered bacterial community composition, which in turn relied on antibiotic administration.

$$ A\rightarrow B\rightarrow C $$

In addition to bacterial communities, our findings highlight a strong link between gut virus communities and colorectal cancer. The signal is dominated by bacteriophages, suggesting the virome may indirectly promote cancer by modulating bacterial community composition within the gut. These results have led us to suggest an alternative model for microbiome-driven colorectal cancer that depends on bacteriophages (*P*), instead of antibiotics, as community modulators.

$$ P\rightarrow B\rightarrow C $$

Our work supports a model in which bacteriophage communities play a role in colorectal cancer progression by altering bacterial community composition, which is already known to impact colorectal cancer. Alpha and beta diversity of the phage communities were not different between cancer states, but a more sophisticated random forest approach suggested that abundance relationships between phage taxa are highly linked to cancer status. Some select viral genomic units (phages) stood out as being highly important to the cancer-phage link, and this importance was correlated with centrality in the bacteria-phage ecological network. This suggests a model in which the phages most associated with colorectal cancer are also hubs within the bacteria-phage ecological network, thus allowing their abundance changes at the center of the interaction network to ripple throughout the rest of the community and promote the cancer-related ecosystem **(Figure \ref{modelsummary})**.

When interpreting these results, it is worth noting that we are studying these communities as operational units, making us unable to draw conclusions with a high taxonomic resolution. This is especially true for bacteriophages, which are poorly annotated and have taxonomy of limited utility. Bacterial operational taxonomic units often represent relatively large categories of bacterial taxa, such as genera, that can represent a variety of species and strains. Likewise, metagenomic operational genomic units represent operationally informative classifications for understanding the community, but should not be interpreted as single phages or bacterial strains.

Further work will be required to support the potential causal relationships outlined in this study. Although we provide evidence for gut bacteriophage communities to alter bacterial communities toward a cancerous state, we must follow up with experimental validation of these observations. The clear pathway toward these answers will be the utilization of murine models as described by Zackular *et al* which showed the link between antibiotic administration and colorectal cancer development [@Zackular:2016en].

@@

![*Example visualization of the proposed model for bacteriophage modulation of colorectal cancer. A) Example network diagrams illustrating the impact of changes on keystone phages (\* asterisk) within the microbiome. A hypothetical ecological network may being at a healthy equilibrium but an alteration to the abundance of a keystone phage can impact the remainder of the network. Highly connected phage host abundances may be altered by the original disruption, which would ripple out to the remainder of the model. B) Another visualization of the impact a change in keystone phage abundance could have on the remainder of the network.*\label{modelsummary}](../figures/ExampleHypothesis.pdf){ width=75% }

@@

![*Lysogenic phage relative abundance in disease states. Phage OGUs were predicted to be either lytic or lysogenic, and the relative abundance of lysogenic phages was quantified and represented as a boxplot. No disease groups were statistically significant.*\label{betaogunegative}](../figures/lysogenic-relabund.pdf){ width=50% }

@@

![*Relative abundance correlations between bacterial OTUs and virome OGUs. A) Pearson correlation R values between all bacterial OTUs (x-axis) and viral OGUs (y-axis) with blue being positively correlated and red being negatively correlated. Operational units are organized by importance in their colorectal cancer classification models, such that the most important units are in the top left corner. B) Magnification of the boxed region in pannel (A), highlighting the correlation between the most important bacterial OTUs and virome OGUs. The most important operational units are in the top left corner of the heatmap, and the correlation scale is the same as pannel (A). C) Histogram quantifying the frequencies of Pearson correlation coefficients between all bacterial OTUs and virome OGUs.*\label{betaogunegative}](../figures/phage-bacteria-cor.pdf){ width=100% }

>>>>

---
title: Characterizing the Role of the Human Virome in Colorectal Cancer
author: 
- name: Geoffrey D Hannigan
  affiliation: 1
- name: Melissa B Duhaime
  affiliation: 2
- name: Mack T Ruffin IV
  affiliation: 3
- name: Charlie C Koumpouras
  affiliation: 1
- name: Patrick D Schloss
  affiliation: 1,\*
address:
  - code: 1
    address: Department of Microbiology & Immunology, University of Michigan, Ann Arbor, Michigan, 48108
  - code: 2
    address: Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, Michigan, 48108
  - code: 3
    address: Department of Family and Community Medicine, Pennsylvania State University Hershey Medical Center, Hershey, Pennsylvania, 17033
  - code: \*
    address: To whom correspondence may be addressed.
output:
  md_document:
    variant: markdown
---

@@

Colorectal cancer is the second leading cause of cancer-related death in the United States and is a primary cause of morbidity and mortality throughout the world. Although the cause of colorectal cancer remains unclear, it has been strongly linked to colon bacterial communities. Viruses are another important component of the colon microbial community that have yet to be studied in colorectal cancer, despite their oncogenic potential. We evaluated the colon virome (virus community) role in colorectal cancer using a cohort of 90 human subjects with either healthy, precancerous, or cancerous colons. We utilized 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic methods to compare the virome's role to that of the bacterial community. We found that alpha and beta diversity metrics were insufficient for detecting an association between the virome and colorectal cancer, but virome-based classification models were highly associated with colorectal cancer. Bacteriophages, not eukaryotic viruses, made up the majority of the CRC-associated virome, suggesting the community was indirectly linked to colorectal cancer, modulating bacterial community structure and functionality. Phages with broader host ranges had a more significant role in cancer development. These results provide foundational evidence that phage communities are associated with colorectal cancer, and that broadly tropic phages play more significant roles. Because of its importance, the virome should be considered in future colorectal cancer studies, as well as other cancer types.

@@

# Significance Statement
Colorectal cancer is a leading cause of cancer-related death in the United States and worldwide. It's progression and severity is linked to colon bacterial communities. Little is known about the cancer-associated colon virus communities and their influence on bacteria. We characterize this influence of colon virus community structure on colorectal cancer in humans. The link between the colon virome and colorectal cancer was established by using virome-based machine learning models to accurately classify patient cancer status. The cancer-virus link was driven primarily by bacteriophages (bacterial viruses). Wider phage host ranges were more strongly linked to colorectal cancer. The results suggest an indirect role for the virome impacting colorectal cancer by modulating their associated bacterial community.

@@

Development of colorectal cancer is a stepwise process that begins when healthy tissue develops into a precancerous polyp (i.e. adenoma) in the large intestine [@Fearon:2011go]. If left untreated, the adenoma will develop into a cancerous lesion that can invade and metastasize, leading to severe illness and death. Progression to cancer can be prevented when precancerous adenomas are detected and removed during routine screening [@Levin:2008kz; @Zauber:2015cr]. Survival for colorectal cancer patients may exceed 90% when the lesions are detected early and removed [@Levin:2008kz]. Screening methods are effective, but their invasiveness has created a lack of compliance, creating a need for accurate, non-invasive screening methods. One such method is screening associated colon microbial communities.

Although the cause of colorectal cancer remains unclear, it has been strongly associated with colon bacterial communities [@Flynn:2016iu; @Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb]. This association has allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja]. While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and understanding underlying etiology, bacteria are only a subset of the colon microbiome. Viruses are another important component of the colon microbial community that have yet to be studied in the context of colorectal cancer.

Due to their mutagenic abilities and their propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Additionally, because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk], and because bacteria have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb], bacteriophages are potentially indirectly linked to cancer. The colon virome (the virus community of the colon) has the potential to impact health and disease (e.g. cancer), and has been associated with diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb]. We aim to take this field of research further by beginning to assess the role of the virome in colorectal cancer.

@@

## Machine Learning Links the Virome to Colorectal Cancer
Previous work has shown that 16S rRNA community signatures are effective for classifying stool samples as originating from healthy, precancerous, or cancerous individuals [@Zackular:2014fba; @Baxter:2016dja]. This is valuable because it presents a potential alternative screening approach to an invasive and expensive colonoscopy. The exceptional performance of bacteria in these classification models also supports a role for bacteria in colorectal cancer. Here we built off of these findings by evaluating the ability of virus community signatures to classify stool samples and compared performance to models built using bacterial community signatures.

@@

We compared the bacterial 16S rRNA gene model to a model built using the virome signatures. The viral model performed as well as the bacterial model (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) with the viral and bacterial models achieving mean AUC (area under the curve) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = sigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = sigfig)`, respectively, after fifteen random forest iterations **(Figure \ref{predmodel} A - B)**.

To confirm that this observation was due to the virome community itself, and was not a trait of metagenomic datasets in general, we built a model using whole metagenomic community signatures. This model performed poorly with a mean AUC of `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = sigfig)`, lending support to our observation that classification performance is a trait unique to the virus metagenome **(Figure \ref{predmodel} A - B)**. Further investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and lowly abundant OTUs. Further filtration of OTUs with a median abundance of zero resulted in the removal of six OTUs, and a loss of model performance down to what was observed in the metagenome **(Figure \ref{16scompare} A)**. The majority of these OTUs had a relative abundance lower than 1% **(Figure \ref{16scompare} B)**.

@@

## The colon Virome Allows for Cancerous, Adenomatous, and Healthy Classification
After evaluating our ability to classify samples as cancerous or healthy, we incorporated the precancerous adenoma samples into the model and evaluated our ability to classify all three states in our total sample set. We used three-class random forest models for the bacterial 16S and viral sample sets. The bacterial signature model yielded an AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = sigfig)` and outperformed the viral community model which yielded an overall AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = sigfig)` (p-value = `r signif(threesig, digits = sigfig)`, **Figure \ref{threewaymodel} A-C**). Both models were best able to classify cancer samples from healthy or precancerous samples, but struggled to distinguish precancerous from healthy or cancerous **(Figure \ref{threewaymodel} A-B)**. The cancerous signal was the most discriminatory of the three sample types.

@@

## Temperate Bacteriophages Drive Link Between Virome and Colorectal Cancer
The virome-based model was able to accurately classify stool samples as cancerous, precancerous, or healthy. Not only is this important for establishing an alternative diagnostic model, but it also suggests an underlying biological importance for viruses in colorectal cancer. Above we used our classification models to evaluate which virus OGUs were most highly linked to colorectal cancer. We were able to further characterize these important OGUs to better understand the underlying biological link between the virome and colorectal cancer.

The role of the virome in colorectal cancer could have been driven directly by eukaryotic viruses or indirectly by bacteriophages acting through their bacterial hosts. To better understand the types of viruses that are important for colorectal cancer, we utilized the longest sequences from the OGUs as the representative sequences to be used for taxonomic classification. These sequences were aligned to a set of all reference virus genomes, including bacteriophages and eukaryotic viruses. A strict e-value threshold of 1e-25 was used to improve our confidence in the matches between the genome sequences. The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel} E)**. Overall we were able to identify `r signif(idcount, digits = sigfig)`% of the OGUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OGUs aligned to bacteriophage reference genomes.

We evaluated whether the phages in the community were primarily lytic (lyse their hosts) or temperate (lysogenic; can integrate into their host's genome, as well as lyse the cell). We accomplished this by identifying three markers for temperate phages in the OGU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes. This approach was done as described in previous work [@Minot:2011ez; @Hannigan:2015fz]. We found that the majority of the colon phages were temperate, and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{threewaymodel} E)**. Thus the majority of the OGUs are temperate bacteriophages and not eukaryotic viruses, indicating the association between the virome and colorectal cancer is reliant on bacteriophage communities that can lie dormant in bacterial genomes. These findings are consistent with previous reports suggesting the colon virome is primarily temperate phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb; @Reyes:2010cwa].

## Influential Bacteria and Phages Associate Indirectly
Because the link between colorectal cancer and the virome was driven by bacteriophages, we hypothesized that the influential phages were predators of the influential bacteria, and thus influenced their relative abundance through predation (e.g. One of the important phages targets *Fusobacterium*). If this hypothesis was true, we would expect that the relative abundance between many important bacteria and phages to be correlated. We correlated the relative abundance values of every phage OGU and bacterial OTU and found strikingly low correlation between the bacterial and phage abundances **(Figure \ref{correlations} A,C)**. There was an overall absence of correlations between the most influential phage OGUs and bacterial OTUs **(Figure \ref{correlations} B)**. This evidence supported our null hypothesis that the influential phages are not predators of the influential bacterial.

We further investigated the infectious capabilities and potential host ranges of the influential phages by building a machine learning model to predict which phage OGUs infect which bacteria in the overall community. The predicted interactions were then used to build a network which was used for subsequent analysis. This approach was done as has been previously described (**Cite Network Preprint**). This analysis revealed a wide tropism range for the bacteriophages within the community **(Figure \ref{network} A)**. We calculated the alpha centrality (measure of importance in the ecosystem network) of each phage OGU's connection to the rest of the network, and compared the centrality to the importance of each OGU in the colorectal cancer classification model. We found that phage OGU centrality is significantly positively correlated with importance to the disease model (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, rho = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), indicating that phages important in driving colorectal cancer are also community hubs **(Figure \ref{network} B)**.

## Influential Phage Identities Shift During CRC Progression
Because our cohort sampled healthy, adenomatous, and cancerous colons, we are able to gain initial insight into virus community shifts during cancer progression. We evaluated the community changes at the two disease transitions (healthy to adenomatous and adenomatous to cancerous) by building individual random forest models to compare only those sample classes. We found that, while bacterial 16S rRNA models perform equally well for all disease class comparisons, the virome models performed worst when seperating healthy from adenomatous, better when seperating adenomatous from cancerous, and the best when seperating healthy from cancerous **(Figure \ref{transitionmodels} A-B)**. Like bacterial communities, the most important phage OGUs provide insight into which organisms are driving the transitions between disease states. Like bacteria **(Figure \ref{transitionmodels} F-H)**, different virome members drove the transitions from healthy to adenomatous and adenomatous to cancerous, with one phage exception **(Figure \ref{transitionmodels} C-E)**. A *Myoviridae* OGU (Cluster 188) was the seventh most important phage driving the transition from a healthy to adenomatous state, and was the most important phage driving the progression from an adenomatous to carcingenic state. Therefore, like bacteria, there are distinct phages assocaited with the transitions from a healthy to an adenomatous colon, and from an adenomatous to a cancerous colon.

# Discussion
Because of their propensity for mutagenesis and capacity for modulating their host functionality, many viruses are oncogenic [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Because some bacteria also have oncogenic properties, bacteriophages may play an indirect role in promoting carcinogenesis by altering bacterial communities [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb]. Despite their carcinogenic potential and the strong association between bacteria and colorectal cancer, the link between virus colon communities (the human colon virome) and colorectal cancer has yet to be evaluated. Here we show that, like colon bacterial communities, the colon virome is associated with colorectal cancer. Our findings support a working hypothesis for oncogenesis by phage-modulated bacterial community composition.

We found that the basic diversity metrics of alpha diversity (richness and Shannon entropy) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community changes between healthy and cancerous states. This led us to implement a more sophisticated machine learning approach (random forest classificaiton model) that allowed us to detect strong associations between the colon virus communities and colorectal cancer. Unlike other human cancer types that are often assocaited with human viruses, colorectal cancer is primarily assocaited with altered bacteriophage communities. These phage communities were not predators of the most influential bacteria (e.g. *Fusobacterium* linked to decreases in *Fusobacterium* phages), as demonstrated by the lack of correlation between the abundance of these entities. Additionally, we identified a correlation in which those phages strongly linked to colorectal cancer were also identified as influential hubs within the overall phage-bacteria network, linking the influence of a phage on the community as a whole to its importance to cancer progression. Our previous work has shown that modifying colon bacterial communities with antibiotics alters colorectal cancer progression and tumor burden in mice, and suggests phage-mediated community alterations may have a similar effect, especially if those phages were broadly infectious [@Zackular:2016en]. Together this data supports a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**.

To better understand the potential mechanisms by which phages modulate the colon bacterial community, we built a working hypothesis that incorporated our findings into the Flynn model for colorectal cancer developement **(Figure \ref{modelsummary} B)** [@Flynn:2016iu]. We hypothesized that the broadly infectious phages in the colon began lysing, and thereby disrupting, the bacterial communities to open a niche in which opportunistic bacteria were able to colonize, such as *Fusobacterium nucleatum*. Once the "driver" bacteria had established themselves in the epithelium, other opporunistic "passenger" bacteria were able to adhere to the driver, colonize, and begin establishing a biofilm. We feel it is likely that the phages played a role in biofilm dispersal and growth, by lysing bacteria within the biofilm, a process shown to be important for effective biofilm growth. The oncogenic bacteria were then able to transform the epithelial cells and disrupt tight junctions to infiltrate the epithelium, thereby initiating an inflammatory immune response. As the adenomatous polyps developed and progressed towards carcinogenesis, we observed a shift in the phages and bacteria important to our cancer classification model. As the bacteria entered their oncogenic synergy with the epithelium, we hypothesize the phages could continue mediating biofilm dispersal, as well as support the colonized oncogenic bacteria by lysing competing cells to maintain the niche and provide nutrients to the other bacteria. In addition to highlighting the most likely mechanisms by which the colorectal cancer virome is interacting with the bacterial communities, this outline will guide the future, mostly functional studies, by our group and others, into the role the virome plays in colorectal cancer.

Although we have evidence for the interaction between the bacteriophage and bacterial community, we are not able to address whether the phages are directly interacting with the human host **(Figure \ref{modelsummary} A)**. Like most foreign entities, bacteriophages are known to be capable of acting as human antigens and thereby initiate or contribute to an immune response. In the context of colorectal cancer, some phages may have been induced out of their bacterial host (most phages were temperate) and interacted with the human host to promote the cancer-associated inflammatory response. Because we are not able to address this behavior with our dataset, we noted it as a known unknown, and another plausible mechanism by which some phages could be associated with colorectal cancer developement.

A notable observation from our analysis was the lack of performance observed using bacterial metagenomic methods compared to the performance of models using viral metagenomes or 16S rRNA gene sequences. This observation highlights the importance of high sequencing coverage in bacterial metagenomic studies, and the advantage of 16S rRNA over whole metagenomic shotgun sequencing. We found that there were six bacterial OTUs that drove the performance of the 16S rRNA classiciation model, and these OTUs were all sparsely present and lowly abundant. Filtration of OTUs with a median relative abundance of zero resulted in the removal of the six important OTUs and reduced model performance to being nearly random like the bacterial metagenomic model. The bacterial metagenomic OGUs represented only the most abundant taxa, which was uninformative for this application. There has been some success in using shotgun metagenomic approaches for stool colorectal cancer classification, but these approaches did not utilize OGU clustering like we did here, and the models only performed **as well** as the 16S rRNA model. Thus the targeted 16S rRNA sequencing approach, which yielded only a fraction of the bacterial metagenomic sequences, was more effective for detecting colorectal cancer in stool samples. Despite a loss of enthusiasm in favor of shotgun metagenomic techniques, 16S rRNA gene sequencing is still a superior methodological approach for some important applications.

@@

![*Working hypothesis of how the bacteriophage community is associated with colorectal cancer and the associated bacterial community.*\label{modelsummary}](../figures/WorkingHypothesis.pdf){ width=100% }

@@

![*Lysogenic phage relative abundance in disease states. Phage OGUs were predicted to be either lytic or lysogenic, and the relative abundance of lysogenic phages was quantified and represented as a boxplot. No disease groups were statistically significant.*\label{replicationstyles}](../figures/lysogenic-relabund.pdf){ width=50% }

@@

![*Relative abundance correlations between bacterial OTUs and virome OGUs. A) Pearson correlation coefficient values between all bacterial OTUs (x-axis) and viral OGUs (y-axis) with blue being positively correlated and red being negatively correlated. Operational units are organized by importance in their colorectal cancer classification models, such that the most important units are in the top left corner. B) Magnification of the boxed region in pannel (A), highlighting the correlation between the most important bacterial OTUs and virome OGUs. The most important operational units are in the top left corner of the heatmap, and the correlation scale is the same as pannel (A). C) Histogram quantifying the frequencies of Pearson correlation coefficients between all bacterial OTUs and virome OGUs.*\label{correlations}](../figures/phage-bacteria-cor.pdf){ width=100% }

![*Comparison of bacterial 16S rRNA classification models with and without OTUs whose median relative abundance are greater than zero. A) Classification model performance (measured as area under the curve) for bacteria models using 16S rRNA data both with and without filtering of samples whose median was zero. Significance was calculated using a Wilcoxon rank sum test, and the resulting p-value is shown. The random area under the curve (0.5) is marked with a dashed line. B) Relative abundance of the six bacterial OTUs removed when filtered for OTUs with median relative abundance of zero. OTU relative abundance is seperated by healthy (red) and cancerous (grey) samples. Relative abundance of 1% is marked by the dashed line.*\label{16scompare}](../figures/filtered16S.pdf){ width=100% }

![*Transition of colorectal cancer importance through disease progression.*\label{transitionmodels}](../figures/temporalmodels.pdf){ width=80% }
